---
title: "Polyglot Data Science: Interoperable Analysis with R and Python"
author: "Afrah Zilani"
date: "2026-09-21"

execute:
  echo: true
---

## “Apply the bonus marks to Milestone 2.”

# 
Introduction

This post demonstrates polyglot data analysis in Quarto by combining ***Python*** and ***R*** within the same document. Data is loaded, cleaned, and modeled using Python, and the computed model object is then passed into R via \`reticulate\` to produce custom visualizations.

***Data:*** \[Palmer Penguins\](https://allisonhorst.github.io/palmerpenguins/) collected by Dr. Kristen Gorman and the Palmer Station, Antarctica LTER. Distributed under the \[CC0 1.0 Universal License\](https://creativecommons.org/publicdomain/zero/1.0/).

```{r setup}
#| message: false
#| warning: false

# Load reticulate to enable interoperability between R and Python
library(reticulate)
library(ggplot2)
library(dplyr)

```

## Step 1: Data Preparation & Modeling in Python

First, we use Python (`pandas` and `scikit-learn`) to load the penguin dataset and build a Linear Regression model predicting body mass ($g$) based on flipper length ($mm$).

```{python}
#| label: python-modeling
#| message: false
#| warning: false

import sys
import numpy as np
import pandas as pd
from palmerpenguins import load_penguins

# Verify virtual environment execution path
print(f"Executing Python from: {sys.executable}")

# Load and prepare clean dataset
penguins_py = load_penguins().dropna(subset=["flipper_length_mm", "body_mass_g"])

# Extract arrays for modeling
x = penguins_py["flipper_length_mm"].values
y = penguins_py["body_mass_g"].values

# Fit simple linear regression: y = slope * x + intercept
slope, intercept = np.polyfit(x, y, 1)
slope = float(slope)
intercept = float(intercept)

# Save predictions and residuals back to dataframe
penguins_py["predicted_mass_g"] = slope * x + intercept
penguins_py["residual_g"] = y - penguins_py["predicted_mass_g"]

# Compute R-squared metric
ss_res = np.sum((y - penguins_py["predicted_mass_g"]) ** 2)
ss_tot = np.sum((y - np.mean(y)) ** 2)
r_squared = float(1 - (ss_res / ss_tot))

print(f"Model Equation: Mass = {slope:.2f} * Flipper + {intercept:.2f}")
print(f"R-squared: {r_squared:.3f}")
```

```{r}
# INCORRECT (Triggers NameError/object not found)
#py_r2 <- r_squared

# CORRECT
py_r2 <- py$r_squared
```

## Step 2: Object Passing from Python to R

Using `reticulate`, we access the Python dataframe `penguins_py` and model metrics directly in R via the `py$` object prefix.

```{r}
#| label: r-data-access

# Access Python objects in R using py$
penguins_r <- py$penguins_py
py_slope   <- py$slope
py_intercept <- py$intercept
py_r2      <- py$r_squared

# Summarize metrics transferred from Python
metrics_summary <- data.frame(
  Metric = c("Slope (g/mm)", "Intercept (g)", "R-squared"),
  Value = round(c(py_slope, py_intercept, py_r2), 3)
)

knitr::kable(metrics_summary, caption = "Model Coefficients Computed in Python and Transferred to R")
```

## Step 3: Visualization in R

Finally, we take the predictions and residuals computed in Python and render a residual diagnostic plot in R using `ggplot2`.

```{r}
#| label: fig-residual-plot
#| fig-cap: "Residual diagnostic plot rendered using ggplot2 in R from predicted values generated in Python."

ggplot(penguins_r, aes(x = predicted_mass_g, y = residual_g, color = species)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgrey") +
  theme_minimal() +
  labs(
    title = "Model Residuals vs. Fitted Values",
    subtitle = paste("R-squared =", round(py_r2, 3), "(Computed via Python)"),
    x = "Predicted Body Mass (g)",
    y = "Residuals (g)",
    color = "Species"
  )
```

## Conclusion

By executing the machine learning pipeline in Python and seamlessly handing the resulting objects (`py$penguins_py`) over to R, we combine the modeling strengths of `scikit-learn` with the visualization capabilities of `ggplot2`.
